### 2018-03-29 reports/thesis/chapters/5_inference.tex
Completed introduction to empirical null.

The method will be described here. Suppose the following test statistics were obtained $z_1,z_2,\dotdotdot,z_n$. First the histogram needs to be smoothed. \cite{efron2004large} recommended using splines but a kernel density estimate \citep{parzen1962on, friedman2001elements} will be used here because it has simpler mathematical properties. The kernel density estimate is
\begin{equation}
	\widehat{p}(z)=
	\frac{1}{nh}
	\sum_{i=1}^n\phi\left(
		\dfrac{z_i-z}{h}
	\right)
\end{equation}
where $h$ is the bandwidth. The bandwidth was chosen such that
\begin{equation}
	h = 0.9n^{-1/5}\times\text{min}\left(s_z,\text{iqr}_z/1.34\right) + 0.16
\end{equation}
where $s_z$ and $\text{iqr}_z$ are the standard deviation and interquartile range. The parameters of the empirical null are estimated using the kernel density estimate. The resulting estimated density is shown in Figure \ref{fig:inference_sub_density_estimate}.

@@

\subsection{Statistical Justification}

>>>>

The method will be described here. Suppose the following test statistics were obtained $z_1,z_2,\dotdotdot,z_n$. First the histogram needs to be smoothed. \cite{efron2004large} recommended using splines but a kernel density estimate \citep{parzen1962on, friedman2001elements} will be used here because it has simpler mathematical properties. The kernel density estimate is
\begin{equation}
	\widehat{p}_Z(z)=
	\frac{1}{nh}
	\sum_{i=1}^n\phi\left(
		\dfrac{z_i-z}{h}
	\right)
\end{equation}
where $h$ is the bandwidth. The bandwidth was chosen such that
\begin{equation}
	h = 0.9n^{-1/5}\times\text{min}\left(s_z,\text{iqr}_z/1.34\right) + 0.16
\end{equation}
where $s_z$ and $\text{iqr}_z$ are the standard deviation and interquartile range. The parameters of the empirical null are estimated using the kernel density estimate. The resulting estimated density is shown in Figure \ref{fig:inference_sub_density_estimate}.

@@

\subsection{More on the Empirical Null}

\cite{efron2004large} invented the empirical null to tackle problems observed in large scale multiple hypothesis testing, for example in microarrays \citep{hedenfalk2001gene, efron2002empirical, efron2003robbins}. These problems are similar to the one observed at the start of this chapter. The empirical null is now used widely, including in neuroimaging \citep{schwartzman2008false,schwartzman2009empirical}.

For formality, the null and alternative hypotheses shall be described here. Assume that the $z$ statistics under the null hypothesis are such that
\begin{equation}
	Z_i|H_0\sim\normal(\mu,\widehat{\sigma}_0^2)
\end{equation}
for $i=1,2,\dotdotdot,n$ and i.i.d. The hypotheses are then
\begin{equation}
	H_0:\mu = \widehat{\mu}_0
\end{equation}
and
\begin{equation}
	H_1:\mu\neq\widehat{\mu}_0
	\ .
\end{equation}
Essentially null $z$ statistics are assumed to be Normal distributed, with parameters using the empirical null. Statistics which deviates too much from $\widehat{\mu}_0$ are considered to be positive.

The distribution of the $z$ statistics can be described using a mixture model. Let the random variable $Z$ have the probability density function
\begin{equation}
	p_Z(z) =
	\pi_0 p_{Z|H_0}(z) + \pi_1 p_{Z|H_1}(z)
\end{equation}
where $0\leqslant\pi_0\leqslant 1$ and  $\pi_1 = 1-\pi_0$. In addition, the null distribution is
\begin{equation}
	p_{Z|H_0}(z) = 
	\dfrac{1}{\sqrt{2\pi}\widehat{\sigma}_0}
	\exp\left[
		-\dfrac{1}{2}
		\left(
			\dfrac{z-\widehat{\mu}_0}{\widehat{\sigma}_0}
		\right)^2
	\right]
\end{equation}
as it was assumed to be Normal. The non-null distribution $p_{Z|H_1}(z)$ does not need to be specified.

The empirical null parameters can be obtained from the following. Assume that $\pi_0$ is large, say bigger than 0.9 \citep{efron2004large}, and at and around the mode the probability density function is dominated by the null distribution. This implies that
\begin{equation}
	p_{Z}(z) \approx \pi_0 p_{Z|H_0}(z)
\end{equation}
for values of $z$ around the mode. Finding the maximum for $p_{Z}(z)$ and $p_{Z|H_0}(z)$ should yield the same solution. This justifies the use of the mode for the empirical null mean \citep{efron2004large}
\begin{equation}
	\widehat{\mu}_0 = \argmax\widehat{p}_Z(z) \ .
\end{equation} 
The empirical null standard deviation can be obtained by noting that
\begin{equation}
	p_{Z}(z) = 
	\dfrac{\pi_0}{\sqrt{2\pi}\widehat{\sigma}_0}
	\exp\left[
		-\dfrac{1}{2}
		\left(
			\dfrac{z-\widehat{\mu}_0}{\widehat{\sigma}_0}
		\right)^2
	\right] \ .
\end{equation}
Taking the log of both sides
\begin{equation}
	\ln p_{Z}(z) = 
	\ln\left[
		\dfrac{\pi_0}{\sqrt{2\pi}\widehat{\sigma}_0}
	\right]
	-\dfrac{1}{2}
	\left(
		\dfrac{z-\widehat{\mu}_0}{\widehat{\sigma}_0}
	\right)^2
	\ .
\end{equation}
Taking derivatives of the log density
\begin{equation}
	\dfrac{\partial}{\partial z} \ln p_{Z}(z) =
	-\left(
		\dfrac{
			z-\widehat{\mu}_0
		}
		{
			\widehat{\sigma}_0^2
		}
	\right)
\end{equation}
\begin{equation}
	\dfrac{\partial^2}{\partial^2 z} \ln p_{Z}(z) =
	-\dfrac{
		1
	}
	{
		\widehat{\sigma}_0^2
	}
\end{equation}
which motivates the estimator \citep{efron2004large}
\begin{equation}
	\widehat{\sigma}_0 = \left[
		\left.
			-\dfrac{\partial^2}{\partial z^2}\ln\widehat{p}(z)
		\right|_{z=\widehat{\mu}_0}
	\right]^{-1/2}
	\ .
\end{equation}
The evaluation of $z=\widehat{\mu}_0$ is used as in that region, one would hope that null test statistics would dominate and non-null test statistics would not contribute much to the density estimate.

This method does not need estimations of $\pi_0$ which makes it quite useful. The estimations of $\pi_0$ is discussed in literature such as \cite{benjamini2000adaptive, pounds2003estimating, storey2003statistical, pounds2004improving, langaas2005estimating, durnez2014posthoc}. Because of the assumptions made on the empirical null, the easiest method is
\begin{equation}
	\widehat{\pi}_0 = \dfrac{\widehat{p}_Z(z)}{p_{Z|H_0}(\widehat{\mu}_0)}
	\ .
\end{equation}
\cite{efron2004large} also motives the use of the empirical null by investigating various types of false discovery rates \citep{storey2002direct, storey2003positive, efron2002empirical, efron2007size} which will not be discussed here.
